Computing in the Field of Rationals

### 2012-07-05 fields.tex
fields

%% The language {{Pi}} however lacks a natural representation for
%% first-class functions due to the first-order nature of its values. In
%% the light of the work of duality of computation it was felt that
%% ``right model'' for expressing functions in {{Pi}} would also express
%% symmetries similar to those in the calculi of Filinski, Curien and
%% Herbelin and Wadler.

The technical goal of this paper is to present the compuational model
{{PiEE}}, a Filinski-style symmetric dual extension of {{Pi}}.

The
extended language, which we call {{PiEE}}, is interesting because it
does not have one De Morgan style dualizing action $(\_)^\bot$, but
two distinct dualizing actions -- an additive duality characterized by
negative types $-(\_)$ and a multiplicative duality characterized by
fractional types $1/(\_)$. Each dualizing operation preserves the
underlying connective {{-(a+b)<->(-a)+(-b)}} and
{{1/(a*b)<->1/a*1/b}}. Negative and fractional types have a clear and
intuitive semantics in the setting of {{Pi}}.

The types of {{PiEE}} correspond to the \emph{field of rational
  numbers} (with the important restriction of fields that there is no
division by zero). Consequently algebraic manipulations of rationals
can be given an operational interpretation and \emph{have
  computational content}.  These negative and fractional dualities
must not confused with $(\-)^\bot$ duality in the additive and
multiplicative fragment of linear logic.  Models of linear logic rely
on *-autonomous categories which change the monoidal tensor when
dualizing. Compact closed categories preserve the monoidal tensor on
dualizing.  The term model of {{PiEE}} corresponds to the
\emph{categorification of a field}. The category is compact closed
over the additive monoid. In the multiplicative monoid (since there is
no division by zero) respects compact closed structure only when a
dual object is defined.

Say more: fractionals, entanglements, backtracking, superposition,
constraints, negative information. Negative types, backward information
flows, sets with negative types, iteration.. etc.

@@

In a recent paper~\cite{infeffects}, we argued that, because they include
irreversible physical primitives, conventional abstract models of computation
have inadvertently included some \emph{implicit} computational effects which
we called \emph{information effects}. We then developed a pure reversible
model of computation that is obtained from the type isomorphisms and
categorical structures that underlie models of linear logic and quantum
computing and that treats information as a linear resource that can neither
be erased nor duplicated. In this paper, we show that our pure reversible
model unveils deeper and more elegant symmetries of computation than have
previously been reported. In particular, we expose two notions of duality of
computation: an additive duality and a multiplicative duality that give rise
to negative types and fractional types respectively. Although these types
have previously appeared in the literature (see Sec.~\ref{sec:related}), they
have typically appeared in the context of conventional languages with
information effects, which limited their appeal and obscured their properties.

@@

>>>>

The language {{Pi}} however lacks a natural representation for
first-class functions due to the first-order nature of its values. The
first approach to try is to re-use Filinski's duality (or one of its
later variants, such duality in the linear logic setting).  The De
Morgan style duality in Filinski's symmetric \lcal (and later works)
is computationally interpreted as follows: if one has a pair of values
{{(v1, v2)}}, then the context that consumes the value has a choice of
comsuming either {{v1}} or {{v2}}. Hence the dual of {{(v1,v2)}} maybe
thought of as the continuation that does {{fst}} or {{snd}}. Such a De
Morgan style computational interpretation is inherently based on the
deletion of values and completely collapses the logical reversibility
and information preservation properties of {{Pi}}. Hence classical De
Morgan duality and linear logic style De Morgan duality (which appears
in later works on polarised focalized $\mu\tilde{\mu}$) do not apply
directly in the context of {{Pi}}.

The other obvious approach to try and build functions is to construct
a ``closed'' category. It is well know that the Int-construction of
Joyal et al \cite{joyal1996traced} and the $\mathcal{G}$ construction
of Abramsky allow the construction of compact closed category
(sometimes also called monoidal closed) from a traced symmetric
monoidal category. The traced extension of {{Pi}}, called {{Pi0}} has
been studied before and indeed the Int-construction gives us a compact
closed structure in this setting. However, the Int-construction does
not preserve the bimonoidal structure of {{Pi}}, i.e. if we use an
additive trace, we lose the {{(1, *)}} monoid and distributivity.

%% In the light of the work of duality of computation it was felt that
%% ``right model'' for expressing functions in {{Pi}} would also
%% express symmetries similar to those in the calculi of Filinski,
%% Curien and Herbelin and Wadler.

The technical goal of this paper is to present the compuational model
{{PiEE}}, a Filinski-style symmetric dual extension of {{Pi}}.  The
extended language has two distinct dualizing actions -- an additive
duality characterized by negative types $-(\_)$ and a multiplicative
duality characterized by fractional types $1/(\_)$. Each dualizing
operation preserves the underlying connective {{-(a+b)<->(-a)+(-b)}}
and {{1/(a*b)<->1/a*1/b}}.  The types of {{PiEE}} correspond to the
\emph{field of rational numbers}. Consequently algebraic manipulations
of rationals can be given an operational interpretation and \emph{have
  computational content}.

%% Negative and fractional types have an
%% intuitive semantics in the setting of {{Pi}}.

The term model of {{PiEE}} corresponds to the \emph{categorification
  of a field}. The category is compact closed over the additive
monoid. In the multiplicative monoid (since there is no division by
zero) respects compact closed structure on non-zero objects.  Compact
closed categories preserve the monoidal tensor on dualizing. In other
words, negative and fractional dualities must not confused with
$(\-)^\bot$ duality in the additive and multiplicative fragment of
linear logic.  Models of linear logic rely on *-autonomous categories
which change the monoidal tensor when dualizing.

%% Say more: fractionals, entanglements, backtracking, superposition,
%% constraints, negative information. Negative types, backward information
%% flows, sets with negative types, iteration.. etc.

@@

% \begin{comment}

%% In a recent paper~\cite{infeffects}, we argued that, because they
%% include irreversible physical primitives, conventional abstract
%% models of computation have inadvertently included some
%% \emph{implicit} computational effects which we called
%% \emph{information effects}. We then developed a pure reversible
%% model of computation that is obtained from the type isomorphisms
%% and categorical structures that underlie models of linear logic and
%% quantum computing and that treats information as a linear resource
%% that can neither be erased nor duplicated. In this paper, we show
%% that our pure reversible model unveils deeper and more elegant
%% symmetries of computation than have previously been reported. In
%% particular, we expose two notions of duality of computation: an
%% additive duality and a multiplicative duality that give rise to
%% negative types and fractional types respectively. Although these
%% types have previously appeared in the literature (see
%% Sec.~\ref{sec:related}), they have typically appeared in the
%% context of conventional languages with information effects, which
%% limited their appeal and obscured their properties.

@@

% \end{comment}
